---
title: "Tracking gender data on COVID-19 series"
author: "Lorenz"
date: "3/13/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

In this blog, we summarize the existing data on sex-disaggregated COVID-19 cases and deaths from [Global Health 50/50](https://globalhealth5050.org/the-sex-gender-and-covid-19-project/the-data-tracker/) and ask how complete our picture is when compared to all reported cases and deaths. In addition, we examine the shares of COVID-19 cases and deaths that are sex-disaggregated by income and region.

```{r import_data, echo=FALSE, warning=FALSE, message=FALSE}
library(httr)
library(jsonlite)
library(tidyverse)

### IMPORT RAW DATA ####

# Set date variables to toggle between versions of data to import.
# Using 2 digit month, 2 digit day format
month <- "03"
day <- "10"

# Import ODW master codes for merging and country groups
odw_master_codes <- read_csv("Input/2021 ODW Country and Region Codes.csv") %>%
  # Clean all variable names by converting to snake case
  janitor::clean_names() %>% 
  # Clear out extra lines at the bottom that just contain notes
  filter(!is.na(country_name)) %>%
  # Clear out duplicate Faroe Islands
  distinct(iso_3_code, .keep_all = TRUE) %>%
  # Keep only relevant indicators and rename for clarity 
  select(iso3c = iso_3_code, country = country_name, odw_region_name, un_code = un_m49_code,
         incgroup = world_bank_income_group_name, lending_cat = world_bank_lending_code_july_2020,
         wbregion = world_bank_all_income_region_names) %>%
  mutate(un_code = as.numeric(un_code), 
         incgroup = fct_relevel(incgroup, "Low income", "Lower middle income", "Upper middle income", "High income"))

# Import Global health 50/50 Sex-disaggregated data tracker file and clean up variable names
# See documentation here: https://api.globalhealth5050.org/docs/
# Get
get_data <- GET("https://api.globalhealth5050.org/api/v1/summary?data=fullvars")
get_data_text <- content(get_data, "text")
get_data_json <- fromJSON(get_data_text, flatten = TRUE)

# Initialize empty dataframe
covid_deaths_cases_raw <- data.frame()
# Run loop through JSON list to extract line of data for every country
# and append to empty dataframe
for (i in 1:length(get_data_json$data)){
  country_df <- as.data.frame(get_data_json$data[i], col.names = "GH5050")
  covid_deaths_cases_raw <- covid_deaths_cases_raw %>%
    bind_rows(country_df)
}


# Import historical Global health 50/50 Sex-disaggregated data tracker file and clean up variable names
# For historical analysis
# Get
get_historical <- GET("https://api.globalhealth5050.org/api/v1/summary?data=historic")
get_historical_text <- content(get_historical, "text")
get_historical_json <- fromJSON(get_historical_text, flatten = TRUE)

# Initialize empty dataframe
gh5050_historical_raw <- data.frame()
# Run loop through JSON list to extract line of data for every country
# and append to empty dataframe
for (i in 1:length(get_historical_json$data)){
  country_df <- as.data.frame(get_historical_json$data[i], col.names = "GH5050")
  gh5050_historical_raw <- gh5050_historical_raw %>%
    bind_rows(country_df)
}

# Import Our World In Data Coronavirus data and clean, keeping date of GH5050 update or
# appending latest date if Gh5050 date isn't available.
# Read in latest data from Our World in Data Github raw
owid_raw <- read_csv("https://raw.githubusercontent.com/owid/covid-19-data/master/public/data/owid-covid-data.csv", guess_max = 20000)
# Below (commented) command preserved December 15 vintage of OWID data to 
# refer to in WDR graph if we need to replicate exactly.
# saveRDS(owid_raw, file = "Input/Our World in Data Dec 15.rds")

owid <- owid_raw %>% 
  # Keep observations for latest Gh5050 update to line up
  filter(date == str_c("2021-", month, "-", day)) %>% 
  # In cases where Gh5050 date is not available (some countries take
  # longer to report, or stopped reporting), append latest date
  # of all countries
  bind_rows(owid_raw %>% 
              group_by(iso_code) %>% 
              mutate(obs_num = row_number()) %>% 
              filter(obs_num == max(obs_num, na.rm = TRUE)) %>% 
              ungroup() %>%
              select(-obs_num)) %>% 
  # Then keep distinct (in this case first) observation by iso3 code 
  # and country name. Will keep Gh5050 update day for countries that 
  # have it and will keep latest update for countries that don't
  # As of October 23, 213 location available, that should be number of obs
  # in dataset
  distinct(iso_code, location, .keep_all = TRUE) %>%
  # Keep only variables we want
  select(iso3c = iso_code, country = location, date, total_cases, total_deaths) %>%
  # replace iso3 value for Kosovo with version that World Bank uses for merging.
  mutate(iso3c = case_when(
    country == "Kosovo" ~ "XKX",
    TRUE ~ iso3c
  )) %>%
  # Add country grouping info
  left_join(odw_master_codes, by = c("iso3c"))

# This can now be accessed via API as well.
# # Import countries that have sex AND age disaggregation for cases and deaths
# # Also from Global Health 5050
# covid_age_sex <- read_csv("Input/gh_sex_age_Jul24.csv") %>%
#   filter(country != "Scotland") %>%
#   mutate(
#     country = case_when(
#       country == "England" ~ "United Kingdom",
#       TRUE ~ country
#     ),
#     iso3c = countrycode::countrycode(country, "country.name", "iso3c"))

# Import 2019 Population from UN World Population prospects 2019
# This is a filtered CSV of the "Total Population", "All variants" file
# found here https://population.un.org/wpp/Download/Standard/CSV/
# This is filtered for year (2019), as well as Variant (Medium)
pop_2019 <- read_csv("Input/WPP2019_TotalPopulationBySex.csv") %>%
  janitor::clean_names() %>%
  # UN uses iso3 numeric country identifiers. Because we use iso3 character everywhere else,
  # merge in ISO country name list, numeric, and character code from
  # https://www.iso.org/obp/ui/#search click on country codes and search
  # merging in with inner_join also keeps only those entities that are actual
  # countries, since ISO list does not contain aggregates unlike UN WPP
  inner_join(read_csv("Input/iso_name_char_code.csv") %>%
               janitor::clean_names(), by = c("loc_id" = "numeric")) %>%
  # Select only relevant variables
  select(iso_num = loc_id, country = location, pop_male:pop_total, iso3c = alpha_3_code) %>%
  # Because GH 5050 reports constitutent countries from UK separately, create sub-components for
  # England, Wales, Northern Ireland, Scotland, using ONS data from
  # https://www.ons.gov.uk/peoplepopulationandcommunity/populationandmigration/populationestimates/bulletins/annualmidyearpopulationestimates/mid2019estimates
  # Figure 3 population pyramid has 2019 population estimates and
  # sex breakdown.
  add_row(country = c("England", "Wales", "Scotland", "Northern Ireland"),
          iso3c = c("ENG", "WAL", "SCO", "NIR"),
          pop_total = c(56286.961, 3152.879, 5463.300, 1893.667),
          pop_male = c(27827.831, 1554.678, 2663.003, 932.717),
          pop_female = c(28459.130, 1598.201, 2800.297, 960.950)) %>%
  mutate(
  # Convert all population estimates from thousands
  pop_male = pop_male*1000,
  pop_female = pop_female*1000,
  pop_total = pop_total*1000) %>%
  # filter out UK
  filter(iso3c != "GBR")

# Create variables that hold total world population
world_pop <- read_csv("Input/WPP2019_TotalPopulationBySex.csv") %>%
  janitor::clean_names() %>%
  filter(location == "World") %>%
  mutate(pop_total = pop_total*1000, pop_male = pop_male*1000, pop_female = pop_female*1000)

world_pop_total <- world_pop %>% pull(pop_total)
world_pop_male <- world_pop %>% pull(pop_male)
world_pop_female <- world_pop %>% pull(pop_female)

### ADDITIONAL PROCESSING ####

# Creating master dataset out of GH5050 datasets
covid_deaths_cases <- covid_deaths_cases_raw %>%
  # Clean variable names
  rename_with(~str_remove(., "GH5050.")) %>%
  # Additional clean and add relevant indicators
  mutate(
    # Convert date columns to date format
    across(contains("date"), ~lubridate::mdy(.x)),
    # Convert columns with numeric info to numeric (currently in factor, need to convert to character, then numeric)
    across(starts_with(c("tests", "cases", "deaths", "hosp", "icu", "healthcare", "cfr", "tot", "male", "female", "vac")), ~as.numeric(as.character(.x))),
    # Convert remaining factor variables to character variables
    across(where(is.factor), as.character),
    # Add country codes
    iso3c = countrycode::countrycode(country_code, "iso2c", "iso3c"),
    iso3c = case_when(
      # Fix ISO codes from import
      country_code == "SAO" ~ "STP",
      country_code == "SEY" ~ "SYC",
      country_code == "XK" ~ "XKX",
      # Add country codes for nations of UK for easier merging
      country == "England" ~ "ENG",
      country == "Wales" ~ "WAL",
      country == "Scotland" ~ "SCO",
      country == "Northern Ireland" ~ "NIR",
      TRUE ~ iso3c
    ),
    # Add source name
    source_name = "Global Health 50/50",
    # Create own sex-disaggregated status variable to reflect differences in reporting cases and deaths
    # New Gh5050 makes Both equal "country has ever reported data on both cases and deaths at the same time point."
    # For our purposes, not important for the moment, so go with whether
    # Sex-disaggregated data is available for both cases and deaths manually.
    # Use cases_date and deaths_date for this, since if cases or deaths are 0,
    # even though they're disaggregated, the columns for percent_male will be NA, rather than 0.
    disaggregated_status = case_when(
    !is.na(date_cases) & !is.na(date_deaths) ~ "Both",
    !is.na(date_cases) & is.na(date_deaths) ~ "Cases only",
    is.na(date_cases) & !is.na(date_deaths) ~ "Deaths only",
    TRUE ~ "None"
    )) %>%
  # Drop empty observations
  filter(!is.na(iso3c)) %>%
  # Import population from UN WPP, see above
  left_join(pop_2019 %>% select(-c(country))) %>%
  # Add country groups
  left_join(odw_master_codes %>% select(-c(country, un_code))) %>%
  mutate(incgroup = as.character(incgroup),
  incgroup = case_when(
    iso3c == "ENG" ~ "High income",
    iso3c == "WAL" ~ "High income",
    iso3c == "SCO" ~ "High income",
    iso3c == "NIR" ~ "High income",
    TRUE ~ incgroup
  ),
  wbregion = case_when(
    iso3c == "ENG" ~ "Europe & Central Asia",
    iso3c == "WAL" ~ "Europe & Central Asia",
    iso3c == "SCO" ~ "Europe & Central Asia",
    iso3c == "NIR" ~ "Europe & Central Asia",
    TRUE ~ wbregion
  ),
  odw_region_name = case_when(
    iso3c == "ENG" ~ "Northern Europe",
    iso3c == "WAL" ~ "Northern Europe",
    iso3c == "SCO" ~ "Northern Europe",
    iso3c == "NIR" ~ "Northern Europe",
    TRUE ~ odw_region_name
  ),
  lending_cat = case_when(
    iso3c == "ENG" ~ "..",
    iso3c == "WAL" ~ "..",
    iso3c == "SCO" ~ "..",
    iso3c == "NIR" ~ "..",
    TRUE ~ lending_cat
  ))

# Clean up
gh5050_historical <- gh5050_historical_raw %>%
  # Make column names nice by removing GH5050 slug
  rename_with(~str_remove(., "GH5050.")) %>%
  # clean column types
  mutate(
    # Convert date columns to date format
    across(contains("date"), ~lubridate::mdy(.x)),
    # Convert columns with numeric info to numeric (currently in factor, need to convert to character, then numeric)
    across(starts_with(c("tests", "cases", "deaths", "hosp", "icu", "healthcare", "cfr", "tot", "male", "female", "vac")), ~as.numeric(as.character(.x))),
    # Convert remaining factor variables to character variables
    across(where(is.factor), as.character),
    # Add country codes
    iso3c = countrycode::countrycode(country_code, "iso2c", "iso3c"),
    iso3c = case_when(
      # Fix ISO codes from import
      country_code == "SAO" ~ "STP",
      country_code == "SEY" ~ "SYC",
      country_code == "XK" ~ "XKX",
      # Add country codes for nations of UK for easier merging
      country == "England" ~ "ENG",
      country == "Wales" ~ "WAL",
      country == "Scotland" ~ "SCO",
      country == "Northern Ireland" ~ "NIR",
      # API imports character column country_code NA as NA missing
      country == "Namibia" ~ "NAM",
      TRUE ~ iso3c
    )) %>%
  # Drop empty observations
  filter(!is.na(iso3c)) %>%
  # Import population from UN WPP, see above
  left_join(pop_2019 %>% select(-c(country))) %>%
  # Add country groups
  left_join(odw_master_codes %>% select(-c(country, un_code))) %>%
  mutate(incgroup = as.character(incgroup),
         incgroup = case_when(
           iso3c == "ENG" ~ "High income",
           iso3c == "WAL" ~ "High income",
           iso3c == "SCO" ~ "High income",
           iso3c == "NIR" ~ "High income",
           TRUE ~ incgroup
         ),
         wbregion = case_when(
           iso3c == "ENG" ~ "Europe & Central Asia",
           iso3c == "WAL" ~ "Europe & Central Asia",
           iso3c == "SCO" ~ "Europe & Central Asia",
           iso3c == "NIR" ~ "Europe & Central Asia",
           TRUE ~ wbregion
         ),
         odw_region_name = case_when(
           iso3c == "ENG" ~ "Northern Europe",
           iso3c == "WAL" ~ "Northern Europe",
           iso3c == "SCO" ~ "Northern Europe",
           iso3c == "NIR" ~ "Northern Europe",
           TRUE ~ odw_region_name
         ),
         lending_cat = case_when(
           iso3c == "ENG" ~ "..",
           iso3c == "WAL" ~ "..",
           iso3c == "SCO" ~ "..",
           iso3c == "NIR" ~ "..",
           TRUE ~ lending_cat
         ))

```

## Table 1 Sex-disaggregated data on the COVID-19 pandemic

```{r table_1, echo=FALSE, warning=FALSE, message=FALSE}
### Table 1
# Create master table of information for 1. Number of countries that
# have sex-disaggreation or only 1 of each, 2. Number of cases
# 3. case split between male and female, 4. Number of deaths, 5. death
# split between male and female, 6. Share of world population by category

# Covid cases male and female shares by groups
covid_table_groups <- covid_deaths_cases %>% 
  # Collapse to sum number of cases/deaths and sex-disaggregated cases/deaths
  # By group
  # Also create aggregate for population and count how many countries in each group
  group_by(disaggregated_status) %>% 
  summarize(num_cases = sum(cases_total_sum, na.rm = TRUE), 
            total_cases_male = sum(cases_male, na.rm = TRUE), 
            total_cases_female = sum(cases_female, na.rm = TRUE),
            num_deaths = sum(deaths_total_sum, na.rm = TRUE), 
            total_deaths_male = sum(deaths_male, na.rm = TRUE), 
            total_deaths_female = sum(deaths_female, na.rm = TRUE),
            sum_pop = sum(totpop2020*1000, na.rm = TRUE),
            countries = n_distinct(iso3c)) %>% 
  ungroup() %>% 
  # Create shares of sex-disaggregated cases/deaths and share of population
  mutate(pct_male_c = total_cases_male/num_cases, 
         pct_fem_c = total_cases_female/num_cases,
         pct_male_d = total_deaths_male/num_deaths, 
         pct_fem_d = total_deaths_female/num_deaths,
         share_pop = sum_pop/world_pop_total) %>%
  filter(disaggregated_status != "None") %>%
  select(disaggregated_status, countries, num_cases, pct_male_c, pct_fem_c,
         num_deaths, pct_male_d, pct_fem_d, share_pop)

### Create table 1
# Starting with three rows we already have (Both, Cases only, Deaths only)
covid_table_groups %>%
      # Append a row made up of sums and weighted averages
      bind_rows(
        # This row is made up of 3 segments
        bind_cols(
      # A segment of sums: Number of countries with any disaggregation
      # Total number of cases, deaths, and share of world population
          covid_table_groups %>%
            summarize(disaggregated_status = "Total", countries = sum(countries, na.rm = TRUE),
                      num_cases = sum(num_cases, na.rm = TRUE),
                      num_deaths = sum(num_deaths, na.rm = TRUE),
                      share_pop = sum(share_pop, na.rm = TRUE)),
          # A segment of weighted average male and female case prevalence
          covid_table_groups %>%
            filter(!is.nan(pct_male_c)) %>%
          summarize(pct_male_c = weighted.mean(pct_male_c, num_cases),
                    pct_fem_c = weighted.mean(pct_fem_c, num_cases)),
      # A segment of weighted average male and female death prevalence
          covid_table_groups %>%
            filter(!is.nan(pct_male_d)) %>%
        summarize(pct_male_d = weighted.mean(pct_male_d, num_deaths),
              pct_fem_d = weighted.mean(pct_fem_d, num_deaths)))) %>%
  rename(`Level of disaggregation` = disaggregated_status, `Number of countries with data` = countries, `Cases` = num_cases, `Male cases (%)` = pct_male_c, `Female cases (%)` = pct_fem_c, `Deaths` = num_deaths, `Male deaths (%)` = pct_male_d, `Female deaths (%)` = pct_fem_d, `Countries with data as share of world population (%)` = share_pop) %>%
  knitr::kable()
```

Table 1 summarizes the sex-disaggregated data available as of `r day` `r months(as.Date(str_c("2021-", month, "-", day)))` 2021. There were `r covid_table_groups %>% filter(disaggregated_status == "Both") %>% pull(countries)` countries reporting sex-disaggregated data for COVID-19 cases and deaths. (These include four nations of the United Kingdom – England, Wales, Scotland, and Northern Ireland – that are reported separately.) They represent `r round(covid_table_groups %>% filter(disaggregated_status == "Both") %>% pull(share_pop) * 100,0)` percent of the world’s population (2019 data). `r covid_table_groups %>% filter(disaggregated_status == "Deaths only") %>% pull(countries)` countries report sex-disaggregated data for deaths only. `r covid_table_groups %>% filter(disaggregated_status == "Cases only") %>% pull(countries)` more countries report sex-disaggregated data on cases but lack data on deaths. In total, `r round(covid_table_groups %>% summarize(total_pop = sum(share_pop, na.rm = TRUE))*100,0)`% of people now live in a country with at least some sex-disaggregated data on COVID-19. For countries reporting sex-disaggregated data on both cases and deaths, women make up `r round(covid_table_groups %>% filter(disaggregated_status == "Both") %>% pull(pct_fem_c)*100,0)` percent of cases but only `r round(covid_table_groups %>% filter(disaggregated_status == "Both") %>% pull(pct_fem_d)*100,0)` percent of deaths.

Although stories about reported cases and deaths caused by the COVID-19 virus have mainly covered high-income countries, information about the pandemic in low- and lower-middle-income countries is slowly emerging and will be crucial to monitoring the impacts of the pandemic on people now and the 2030 Agenda. As of `r day` `r months(as.Date(str_c("2021-", month, "-", day)))`, `r covid_deaths_cases %>% filter(incgroup == "Low income", disaggregated_status != "None") %>% count()` low-income countries published sex-disaggregated data (`r covid_deaths_cases %>% filter(incgroup == "Low income", disaggregated_status == "Cases only") %>% count()` of them for cases only and `r covid_deaths_cases %>% filter(incgroup == "Low income", disaggregated_status == "Both") %>% count()` for both cases and deaths). Sex-disaggregated data are available for `r covid_deaths_cases %>% filter(incgroup == "Lower middle income", disaggregated_status != "None") %>% count()` lower-middle-income countries (For FY2021, the World Bank designates `r odw_master_codes %>% filter(incgroup == "Low income") %>% count()` countries as Low income and `r odw_master_codes %>% filter(incgroup == "Lower middle income") %>% count()` countries as Lower middle income). See Table 3 below for a more in-depth discussion on the reporting of deaths and cases by income group.

## Table 2 Global data on COVID-19 cases and deaths
```{r table_2, echo=FALSE, warning=FALSE, message=FALSE}
### Table 2
#### This drops French Polynesia because OWID does not have data on that island
#### How to address not quite independent countries? Including Jersey, for example

# Merge OWID data with list of GH5050 countries and their sex-disaggregation status
owid_working <- owid %>%
  # Take out World aggregate, regional aggregates, "International", which is no longer used (was for cruise ships, etc.)
  filter(!str_detect(iso3c, "OWID")) %>%
  # Clean country name variables
  select(-country.y) %>%
  rename(country = country.x) %>%
  # Merge in GH5050 countries and sex-disaggregation status
  left_join(covid_deaths_cases %>%
              filter(!iso3c %in% c("ENG", "WAL", "SCO", "NIR")) %>%
              add_row(iso3c = "GBR", country = "United Kingdom", disaggregated_status = "Both") %>%
              select(iso3c, disaggregated_status)) %>%
  # Clean sex-disaggregated status indicator
  mutate(disaggregated_status = case_when(
    disaggregated_status == "None" ~ NA_character_,
    TRUE ~ disaggregated_status
  ))

# Create interim df that summarizes number of cases and deaths by
# country groups around sex-disaggregated status: Both disaggregations,
# Cases only, deaths only, and neither.
owid_table <- owid_working %>%
  group_by(disaggregated_status) %>%
    # Note we are keeping cases and deaths for countries that have
    # deaths only/cases only disaggregated instead of dropping them
    # In the blog, we mark them and discuss that these cases/deaths are not
    # actually sex-disaggregated. However, to avoid confusion, we keep them in the same
    # country group line. Need better solution in the future.
  summarize(countries = n_distinct(iso3c), cases_total = sum(total_cases, na.rm = TRUE),
            deaths_total = sum(total_deaths, na.rm = TRUE)) %>%
  ungroup() %>%
  mutate(all_cases = sum(cases_total, na.rm = TRUE), all_deaths = sum(deaths_total, na.rm = TRUE),
         share_cases = cases_total/all_cases, share_deaths = deaths_total/all_deaths) %>%
  select(disaggregated_status, countries, cases_total, share_cases, deaths_total, share_deaths)

# Put in sub-totals of Both, Cases only, and deaths only disaggregations
owid_table_2 <- owid_table %>%
  # Take out the "Neither" line
  filter(!is.na(disaggregated_status)) %>%
  # The function below will put in totals of all the columns, called "Total"
  janitor::adorn_totals(name = "Sub-Total") %>%
  # Append the "Neither" row from the df back
  bind_rows(owid_table %>%
              filter(is.na(disaggregated_status)))

# Add final line that adds sub-total to "Neither" info and creates Global total
owid_table_2 %>%
    # Append total of sub-total and "Neither" category
    bind_rows(owid_table_2 %>%
              # Keep just sub-total and "Neither" group
              filter(is.na(disaggregated_status) | disaggregated_status == "Sub-Total") %>%
                # Use summarize function to create total line 
                summarize(across(countries:share_deaths, sum)) %>%
                # Then add column to match column headers and name it Global Total
                add_column(disaggregated_status = "Global Total", .before = "countries")) %>%
    # Rename "Neither" category to better label
    mutate(disaggregated_status = case_when(
      is.na(disaggregated_status) ~ "Countries w/o sex disaggregation",
      TRUE ~ disaggregated_status
    )) %>% 
  rename(`Level of disaggregation` = disaggregated_status, `Number of countries` = countries, `Total cases reported` = cases_total, `Share of total cases (%)` = share_cases, `Total deaths reported` = deaths_total, `Share of total deaths (%)` = share_deaths) %>%
      knitr::kable()
```

