---
title: "Tracking gender data on COVID-19 series"
author: "Lorenz"
date: "3/13/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

In this blog, we summarize the existing data on sex-disaggregated COVID-19 cases and deaths from [Global Health 50/50](https://globalhealth5050.org/the-sex-gender-and-covid-19-project/the-data-tracker/) and ask how complete our picture is when compared to all reported cases and deaths. In addition, we examine the shares of COVID-19 cases and deaths that are sex-disaggregated by income and region.

```{r import_data, echo=FALSE, warning=FALSE, message=FALSE}
library(httr)
library(jsonlite)
library(tidyverse)

### IMPORT RAW DATA ####

# Set date variables to toggle between versions of data to import.
# Using 2 digit month, 2 digit day format
month <- "03"
day <- "10"

# Import ODW master codes for merging and country groups
odw_master_codes <- read_csv("Input/2021 ODW Country and Region Codes.csv") %>%
  # Clean all variable names by converting to snake case
  janitor::clean_names() %>% 
  # Clear out extra lines at the bottom that just contain notes
  filter(!is.na(country_name)) %>%
  # Clear out duplicate Faroe Islands
  distinct(iso_3_code, .keep_all = TRUE) %>%
  # Keep only relevant indicators and rename for clarity 
  select(iso3c = iso_3_code, country = country_name, odw_region_name, un_code = un_m49_code,
         incgroup = world_bank_income_group_name, lending_cat = world_bank_lending_code_july_2020,
         wbregion = world_bank_all_income_region_names) %>%
  mutate(un_code = as.numeric(un_code), 
         incgroup = fct_relevel(incgroup, "Low income", "Lower middle income", "Upper middle income", "High income"))

# Import Global health 50/50 Sex-disaggregated data tracker file and clean up variable names
# See documentation here: https://api.globalhealth5050.org/docs/
# Get
get_data <- GET("https://api.globalhealth5050.org/api/v1/summary?data=fullvars")
get_data_text <- content(get_data, "text")
get_data_json <- fromJSON(get_data_text, flatten = TRUE)

# Initialize empty dataframe
covid_deaths_cases_raw <- data.frame()
# Run loop through JSON list to extract line of data for every country
# and append to empty dataframe
for (i in 1:length(get_data_json$data)){
  country_df <- as.data.frame(get_data_json$data[i], col.names = "GH5050")
  covid_deaths_cases_raw <- covid_deaths_cases_raw %>%
    bind_rows(country_df)
}


# Import historical Global health 50/50 Sex-disaggregated data tracker file and clean up variable names
# For historical analysis
# Get
get_historical <- GET("https://api.globalhealth5050.org/api/v1/summary?data=historic")
get_historical_text <- content(get_historical, "text")
get_historical_json <- fromJSON(get_historical_text, flatten = TRUE)

# Initialize empty dataframe
gh5050_historical_raw <- data.frame()
# Run loop through JSON list to extract line of data for every country
# and append to empty dataframe
for (i in 1:length(get_historical_json$data)){
  country_df <- as.data.frame(get_historical_json$data[i], col.names = "GH5050")
  gh5050_historical_raw <- gh5050_historical_raw %>%
    bind_rows(country_df)
}

# Import Our World In Data Coronavirus data and clean, keeping date of GH5050 update or
# appending latest date if Gh5050 date isn't available.
# Read in latest data from Our World in Data Github raw
owid_raw <- read_csv("https://raw.githubusercontent.com/owid/covid-19-data/master/public/data/owid-covid-data.csv", guess_max = 20000)
# Below (commented) command preserved December 15 vintage of OWID data to 
# refer to in WDR graph if we need to replicate exactly.
# saveRDS(owid_raw, file = "Input/Our World in Data Dec 15.rds")

owid <- owid_raw %>% 
  # Keep observations for latest Gh5050 update to line up
  filter(date == str_c("2021-", month, "-", day)) %>% 
  # In cases where Gh5050 date is not available (some countries take
  # longer to report, or stopped reporting), append latest date
  # of all countries
  bind_rows(owid_raw %>% 
              group_by(iso_code) %>% 
              mutate(obs_num = row_number()) %>% 
              filter(obs_num == max(obs_num, na.rm = TRUE)) %>% 
              ungroup() %>%
              select(-obs_num)) %>% 
  # Then keep distinct (in this case first) observation by iso3 code 
  # and country name. Will keep Gh5050 update day for countries that 
  # have it and will keep latest update for countries that don't
  # As of October 23, 213 location available, that should be number of obs
  # in dataset
  distinct(iso_code, location, .keep_all = TRUE) %>%
  # Keep only variables we want
  select(iso3c = iso_code, country = location, date, total_cases, total_deaths) %>%
  # replace iso3 value for Kosovo with version that World Bank uses for merging.
  mutate(iso3c = case_when(
    country == "Kosovo" ~ "XKX",
    TRUE ~ iso3c
  )) %>%
  # Add country grouping info
  left_join(odw_master_codes, by = c("iso3c"))

# This can now be accessed via API as well.
# # Import countries that have sex AND age disaggregation for cases and deaths
# # Also from Global Health 5050
# covid_age_sex <- read_csv("Input/gh_sex_age_Jul24.csv") %>%
#   filter(country != "Scotland") %>%
#   mutate(
#     country = case_when(
#       country == "England" ~ "United Kingdom",
#       TRUE ~ country
#     ),
#     iso3c = countrycode::countrycode(country, "country.name", "iso3c"))

# Import 2019 Population from UN World Population prospects 2019
# This is a filtered CSV of the "Total Population", "All variants" file
# found here https://population.un.org/wpp/Download/Standard/CSV/
# This is filtered for year (2019), as well as Variant (Medium)
pop_2019 <- read_csv("Input/WPP2019_TotalPopulationBySex.csv") %>%
  janitor::clean_names() %>%
  # UN uses iso3 numeric country identifiers. Because we use iso3 character everywhere else,
  # merge in ISO country name list, numeric, and character code from
  # https://www.iso.org/obp/ui/#search click on country codes and search
  # merging in with inner_join also keeps only those entities that are actual
  # countries, since ISO list does not contain aggregates unlike UN WPP
  inner_join(read_csv("Input/iso_name_char_code.csv") %>%
               janitor::clean_names(), by = c("loc_id" = "numeric")) %>%
  # Select only relevant variables
  select(iso_num = loc_id, country = location, pop_male:pop_total, iso3c = alpha_3_code) %>%
  # Because GH 5050 reports constitutent countries from UK separately, create sub-components for
  # England, Wales, Northern Ireland, Scotland, using ONS data from
  # https://www.ons.gov.uk/peoplepopulationandcommunity/populationandmigration/populationestimates/bulletins/annualmidyearpopulationestimates/mid2019estimates
  # Figure 3 population pyramid has 2019 population estimates and
  # sex breakdown.
  add_row(country = c("England", "Wales", "Scotland", "Northern Ireland"),
          iso3c = c("ENG", "WAL", "SCO", "NIR"),
          pop_total = c(56286.961, 3152.879, 5463.300, 1893.667),
          pop_male = c(27827.831, 1554.678, 2663.003, 932.717),
          pop_female = c(28459.130, 1598.201, 2800.297, 960.950)) %>%
  mutate(
  # Convert all population estimates from thousands
  pop_male = pop_male*1000,
  pop_female = pop_female*1000,
  pop_total = pop_total*1000) %>%
  # filter out UK
  filter(iso3c != "GBR")

# Create variables that hold total world population
world_pop <- read_csv("Input/WPP2019_TotalPopulationBySex.csv") %>%
  janitor::clean_names() %>%
  filter(location == "World") %>%
  mutate(pop_total = pop_total*1000, pop_male = pop_male*1000, pop_female = pop_female*1000)

world_pop_total <- world_pop %>% pull(pop_total)
world_pop_male <- world_pop %>% pull(pop_male)
world_pop_female <- world_pop %>% pull(pop_female)

### ADDITIONAL PROCESSING ####

# Creating master dataset out of GH5050 datasets
covid_deaths_cases <- covid_deaths_cases_raw %>%
  # Clean variable names
  rename_with(~str_remove(., "GH5050.")) %>%
  # Additional clean and add relevant indicators
  mutate(
    # Convert date columns to date format
    across(contains("date"), ~lubridate::mdy(.x)),
    # Convert columns with numeric info to numeric (currently in factor, need to convert to character, then numeric)
    across(starts_with(c("tests", "cases", "deaths", "hosp", "icu", "healthcare", "cfr", "tot", "male", "female", "vac")), ~as.numeric(as.character(.x))),
    # Convert remaining factor variables to character variables
    across(where(is.factor), as.character),
    # Add country codes
    iso3c = countrycode::countrycode(country_code, "iso2c", "iso3c"),
    iso3c = case_when(
      # Fix ISO codes from import
      country_code == "SAO" ~ "STP",
      country_code == "SEY" ~ "SYC",
      country_code == "XK" ~ "XKX",
      # Add country codes for nations of UK for easier merging
      country == "England" ~ "ENG",
      country == "Wales" ~ "WAL",
      country == "Scotland" ~ "SCO",
      country == "Northern Ireland" ~ "NIR",
      TRUE ~ iso3c
    ),
    # Add source name
    source_name = "Global Health 50/50",
    # Create own sex-disaggregated status variable to reflect differences in reporting cases and deaths
    # New Gh5050 makes Both equal "country has ever reported data on both cases and deaths at the same time point."
    # For our purposes, not important for the moment, so go with whether
    # Sex-disaggregated data is available for both cases and deaths manually.
    # Use cases_date and deaths_date for this, since if cases or deaths are 0,
    # even though they're disaggregated, the columns for percent_male will be NA, rather than 0.
    disaggregated_status = case_when(
    !is.na(date_cases) & !is.na(date_deaths) ~ "Both",
    !is.na(date_cases) & is.na(date_deaths) ~ "Cases only",
    is.na(date_cases) & !is.na(date_deaths) ~ "Deaths only",
    TRUE ~ "None"
    )) %>%
  # Drop empty observations
  filter(!is.na(iso3c)) %>%
  # Import population from UN WPP, see above
  left_join(pop_2019 %>% select(-c(country))) %>%
  # Add country groups
  left_join(odw_master_codes %>% select(-c(country, un_code))) %>%
  mutate(incgroup = as.character(incgroup),
  incgroup = case_when(
    iso3c == "ENG" ~ "High income",
    iso3c == "WAL" ~ "High income",
    iso3c == "SCO" ~ "High income",
    iso3c == "NIR" ~ "High income",
    TRUE ~ incgroup
  ),
  wbregion = case_when(
    iso3c == "ENG" ~ "Europe & Central Asia",
    iso3c == "WAL" ~ "Europe & Central Asia",
    iso3c == "SCO" ~ "Europe & Central Asia",
    iso3c == "NIR" ~ "Europe & Central Asia",
    TRUE ~ wbregion
  ),
  odw_region_name = case_when(
    iso3c == "ENG" ~ "Northern Europe",
    iso3c == "WAL" ~ "Northern Europe",
    iso3c == "SCO" ~ "Northern Europe",
    iso3c == "NIR" ~ "Northern Europe",
    TRUE ~ odw_region_name
  ),
  lending_cat = case_when(
    iso3c == "ENG" ~ "..",
    iso3c == "WAL" ~ "..",
    iso3c == "SCO" ~ "..",
    iso3c == "NIR" ~ "..",
    TRUE ~ lending_cat
  ))

# Clean up
gh5050_historical <- gh5050_historical_raw %>%
  # Make column names nice by removing GH5050 slug
  rename_with(~str_remove(., "GH5050.")) %>%
  # clean column types
  mutate(
    # Convert date columns to date format
    across(contains("date"), ~lubridate::mdy(.x)),
    # Convert columns with numeric info to numeric (currently in factor, need to convert to character, then numeric)
    across(starts_with(c("tests", "cases", "deaths", "hosp", "icu", "healthcare", "cfr", "tot", "male", "female", "vac")), ~as.numeric(as.character(.x))),
    # Convert remaining factor variables to character variables
    across(where(is.factor), as.character),
    # Add country codes
    iso3c = countrycode::countrycode(country_code, "iso2c", "iso3c"),
    iso3c = case_when(
      # Fix ISO codes from import
      country_code == "SAO" ~ "STP",
      country_code == "SEY" ~ "SYC",
      country_code == "XK" ~ "XKX",
      # Add country codes for nations of UK for easier merging
      country == "England" ~ "ENG",
      country == "Wales" ~ "WAL",
      country == "Scotland" ~ "SCO",
      country == "Northern Ireland" ~ "NIR",
      # API imports character column country_code NA as NA missing
      country == "Namibia" ~ "NAM",
      TRUE ~ iso3c
    )) %>%
  # Drop empty observations
  filter(!is.na(iso3c)) %>%
  # Import population from UN WPP, see above
  left_join(pop_2019 %>% select(-c(country))) %>%
  # Add country groups
  left_join(odw_master_codes %>% select(-c(country, un_code))) %>%
  mutate(incgroup = as.character(incgroup),
         incgroup = case_when(
           iso3c == "ENG" ~ "High income",
           iso3c == "WAL" ~ "High income",
           iso3c == "SCO" ~ "High income",
           iso3c == "NIR" ~ "High income",
           TRUE ~ incgroup
         ),
         wbregion = case_when(
           iso3c == "ENG" ~ "Europe & Central Asia",
           iso3c == "WAL" ~ "Europe & Central Asia",
           iso3c == "SCO" ~ "Europe & Central Asia",
           iso3c == "NIR" ~ "Europe & Central Asia",
           TRUE ~ wbregion
         ),
         odw_region_name = case_when(
           iso3c == "ENG" ~ "Northern Europe",
           iso3c == "WAL" ~ "Northern Europe",
           iso3c == "SCO" ~ "Northern Europe",
           iso3c == "NIR" ~ "Northern Europe",
           TRUE ~ odw_region_name
         ),
         lending_cat = case_when(
           iso3c == "ENG" ~ "..",
           iso3c == "WAL" ~ "..",
           iso3c == "SCO" ~ "..",
           iso3c == "NIR" ~ "..",
           TRUE ~ lending_cat
         ))

```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
